For induction-induction, induction is enough

### 2019-09-13 paper/universaliit.tex
Rks Ambrus

The steps consist of the following:
\begin{enumerate}
\item Definition of untyped syntax (as a family of inductive datatypes) together
  with typing judgments (as inductive relations on the untyped
  syntax), and construction of a model of the theory of IIT
  signatures from well-formed terms, denoted $S$.
\item Construction of a morphism $\rec:S\to M$ for arbitrary $M$, by:
  \begin{enumerate}
  \item defining a relation $\blank \sim \blank$ between the syntax and a given
    model (the idea is that given a syntactic context $\Gamma$ and a semantic
    context $\model{\Gamma}$ of the model $M$, we have $\Gamma \sim
    \model{\Gamma}$ if and only if $\rec\,\Gamma = \model{\Gamma}$, and
    similarly for types, terms, and substitutions);
    % semantic context $\model{\Gamma}$ of
    % the model $M$ if and only if $\Gamma$ relates to $\rec\,{\Gamma}$;
    % enjoyed by the initial morphism from the
    % syntax to a given model;
  \item Showing that this relation is functional;
    \end{enumerate}
  \item Proving of the uniqueness of this morphism by showing that
    any morphism $f:S\to M$ satisfies the relation
    (in the sense, for contexts, that given any syntactic context $\Gamma$, we
    have $\Gamma\sim f\,\Gamma$).
      % TODO donner composantes sur les contextes
\end{enumerate}
The next sections detail each of these steps.
\subsection{Syntactic model}

The goal is to define the syntactic model by packing the untyped syntax
with the typing judgments:
% % \subsection{QIIT model}
% Then, we show that they are stable by weakening and substitution.
% Packing the untyped syntax with well-formed typed judgments yields
% a model $S$ of the QIIT. Let us give the most important components:
\begin{alignat*}{5}
 & \syn\Con && := \sum_\Gamma \Gamma\vdash
 \\
 & \syn\Ty \,(\Gamma,\wf\Gamma)&& := \sum_A \Gamma\vdash A
 \\
 & \syn\Tm \,(\Gamma,\wf\Gamma)(A,\wf A)&& := \sum_t \Gamma\vdash t \in A
 \\
 & \syn\Sub \,(\Gamma,\wf\Gamma)(\Delta,\wf \Delta)&& := \sum_\sigma \Gamma\vdash \sigma \Rightarrow \Delta
\end{alignat*}
The next sections are devoted to the definition of the untyped syntax and the
typing judgments.

@@

\begin{alignat*}{5}
  & \rlap{(1) Substitution calculus} \\[0.5em]
  & \Conu && : \Set \\
  & \Tyu  && : \Set \\
  & \Subu  && : \Set \\
  % & \Subpre  && : \Set \\
  & \Tmu  && : \Set \\
  & \emptyu && : \Conu \\
    & \epsilonu && : \Subu \\
  & \blank\extu\blank && : \Conu\ra\Tyu\ra\Conu \\
  & \blank\consu\blank && : \Subu\ra\Tmu\ra\Subu \\
  & \varu  && : \N \ra \Tmu \\
  & \rlap{(2) Sorts and operators} \\[0.5em]
  & \Uu && : \Tyu \\
  & \Elu && : \Tmu\ra\Tyu \\
  & \rlap{(3) Parameters} \\[0.5em]
  & \Piu && : \Tmu\ra\Tyu\ra\Tyu \\
  & \blank\appu\blank && : \Tmu\ra\Tmu\ra \Tmu \\
  & \rlap{(4) Metatheoretic and infinitary parameters} \\[0.5em]
  & \Pimu && : (T:\Set)\ra(T\ra\Tyu)\ra\Tyu \\
  & \Piiu && : (T:\Set)\ra(T\ra\Tmu)\ra\Tmu \\
  & \blank\appmu \blank && : \Tmu\ra(\alpha:T) \ra\Tmu\\
  & \rlap{(5) Default value} \\[0.5em]
  & \erru && : \Tmu
\end{alignat*}
Note that we choose to share the constructor for untyped application to infinitary and
meta-theoretic parameters.
Variables are modelled as De Bruijn indices, i.e. as natural numbers pointing
to the relevant position in the context.

@@

\subsubsection{Weakening preserves typing}

@@

Note that $(\Piu \,A\,B)[\sigma]$ should be defined as $\Piu
\,A[\sigma]\,B[\wk\,\sigma]$, and thus we need to define $\wk$, the weakening substitution.
The basic idea is to increment the de Bruijn indices of all the variables.
Actually this is not so simple because of the $\Piu$ type:
indeed, we want to define $\wk(\Piu\,A\,B)$ as the $\Pi$ type of the weakening
of $A$ and $B$, but here, $B$ must be weakened with respect to the second last
variable of the context, rather than the last one.
For this reason, we need to generalize the weakening as occuring anywhere in the context.
\begin{alignat*}{5}
  & \wk_n && :  \Tyu\ra\Tyu \\
  & \wk_n && :  \Tmu\ra\Tmu \\
  % & \wk && : \N \ra \N\ra\N \\
  & \wkO && : \Subu\ra \Subu
  \end{alignat*}
  The natural number $n$ specifies at which position of the context the
  weakening occurs.
  Here, $\wkO$ weakens with respect to the last variable.

  \subsubsection{Substitution preserves typing}
  We define by induction the specific substitution of one variable of the context:
  % & \rlap{(1) Substitution calculus} \\[0.5em]
\begin{alignat*}{5}
  & \blank[\blank := \blank] && : \Tyu\ra\N\ra\Tmu\ra\Tyu \\
  & \blank[\blank := \blank] && : \Tmu\ra\N\ra\Tmu\ra\Tmu
  % \\
  % & \blank[\blank := \blank] && : \N\ra\N\ra\Tmu\ra\Tmu
  \end{alignat*}
  This is enough to define the well-typed judgments: indeed, the typing rule of application
  involves only a unary substitution.

@@

\subsubsection{Typed weakening}
Untyped weakening performs weakening anywhere in the context.
Stating a typing rule for this operation requires to have a way to express
the weakening occuring at the middle of a context.
To this end, we consider pairs of untyped contexts, which should be thought
of as a splitting at some place of a full context.
This full context is recovered by merging the two components:
% (which appears in the typing rule of a weakening occuring in the middle of a context),
% and the length $\length{\Gamma}$ of a
% context $\Gamma$:
  & \blank \merge \blank && :  \Conu\ra\Conu\ra\Conu \\
  & \Gamma \merge \cdot && :=  \Gamma \\
  & \Gamma \merge (\Delta \extu A) && :=  (\Gamma \merge \Delta)\extu A
We think of the second context as a telescope over the first context.
We need to define the define also the weakening of a telescope in order to
state the typing rule of weakening:
  & \wkO && : \Conu\ra \Conu
  & \wkO\,\emptyu && := \emptyu
  & \wkO\,(\Delta \extu A) && := \wkO\,\Delta \extu \wk_{\length{\Delta}}\,A
Now we state and prove by mutual recursion that typing judgments are stable under
weakening, for contexts, types, terms, and substitutions:
% wkTelw : ∀ {Γp}{Ap}(Aw : Γp ⊢ Ap)Δp (Δw : (Γp ^^ Δp) ⊢) → ((Γp ▶p Ap) ^^ wkTel Δp) ⊢
% liftTw : ∀ {Γp}{Ap}(Aw : Γp ⊢ Ap)Δp{Bp}(Bw : (Γp ^^ Δp) ⊢ Bp) → ((Γp ▶p Ap) ^^ wkTel Δp) ⊢ (liftT ∣ Δp ∣ Bp)
  & \wf{\wkO} && : (\Gamma \vdash A)\ra(\Gamma \merge \Delta\vdash) \ra
  \Gamma \extu A \merge \wkO\, \Delta \vdash \\
  % liftV (S (n + p)) (liftV n q) ≡ liftV n (liftV (n + p) q)
  & \wf{\wk} && : (\Gamma \vdash A)\ra(\Gamma \merge \Delta\vdash B) \ra
  \Gamma \extu A \merge \wkO\,\Delta \vdash \wk_{\length{\Delta}}\, B  \\
  & \wf{\wk} && : (\Gamma \vdash A)\ra(\Gamma \merge \Delta\vdash t\in B) \ra
  \Gamma \extu A \merge \wkO\,\Delta \vdash \wk_{\length{\Delta}}\, t \in \wk_{\length{\Delta}}\,B  \\
  & \wf{\wkO} && : (\Gamma \vdash A)\ra(\Gamma \vdash \sigma
  \Rightarrow \Delta) \ra
  \Gamma \extu A \vdash \wkO\, \sigma \Rightarrow \Delta
\subsubsection{Typed substitution}
  We show that judgments are stable under substitution, by mutual recursion:
  % Tyw[] : ∀ {Γp}{Ap}(Aw : Γp ⊢ Ap) {Δp}(Δw : Δp ⊢){σp}(σw :  Δp ⊢ σp ⇒ Γp) → Δp ⊢ (Ap [ σp ]T)
  & \wf{[]} && : (\Gamma \vdash)\ra(\Delta \vdash A)\ra
  (\Gamma \vdash \sigma \Rightarrow \Delta) \ra
  \Gamma \vdash A [ \sigma ]
  & \wf{[]} && : (\Gamma \vdash)\ra(\Delta \vdash t \in A)\ra
  (\Gamma \vdash \sigma \Rightarrow \Delta) \ra
  \Gamma \vdash t [\sigma] \in A [ \sigma ]
  & \wf{[]} && : (\Delta \vdash x \invar A)\ra
  (\Gamma \vdash \sigma \Rightarrow \Delta) \ra
  \Gamma \vdash x [\sigma] \in A [ \sigma ]
  % ∘w : ∀ {Γ} {Δ σ} (σw :  Δ ⊢ σ ⇒ Γ)
  %  {Y}(Yw : Y ⊢) {δ} (δw :  Y ⊢ δ ⇒ Δ) →
  %  Y ⊢  (σ ∘p δ) ⇒ Γ
  & \wf{{\circ}} && :
  (\Gamma \vdash) \ra
  (\Gamma \vdash \sigma \Rightarrow \Delta) \ra
  (\Delta \vdash \tau \Rightarrow E) \ra
  \Gamma \vdash \tau \circ \sigma \Rightarrow E
  % keepw : ∀ {Γp}(Γw : Γp ⊢){Δp}(Δw : Δp ⊢){σp}(σw :  Γp ⊢ σp ⇒ Δp) {Ap}(Aw : Δp ⊢ Ap ∈ Up ) → (Γp ▶p (Elp Ap [ σp ]T )) ⊢ (keep σp) ⇒ (Δp ▶p Elp Ap)
  % & \wf{keep-\El} && : (\Gamma \vdash)\ra(\Delta \vdash A)\ra
  % (\Gamma \vdash \sigma \Rightarrow \Delta) \ra
  % \Gamma \vdash A [ \sigma ]
Then, we show that identity substitutions are neutral, for well-formed types,
terms, and substitutions:
  % [idp]V : ∀ {Γ}{A}{x}(xw : Γ ⊢ x ∈v A) → (x [ idp ∣ Γ ∣ ]V) ≡ V x
  & [\idu] && : (\Gamma \vdash A)\ra A [ \idu\,\Gamma]  \\
  & [\idu] && : (\Gamma \vdash x \invar A)\ra x [ \idu\,\Gamma] = V x \\
  & [\idu] && : (\Gamma \vdash t \in A)\ra t [ \idu\,\Gamma] = t \\
  & \idru && : (\Gamma \vdash \sigma \Rightarrow\Delta)\ra \sigma \circ \idu\,\Gamma = \sigma
  In a seperate and independant recursion, one shows that:
  & \idlu && : (\Gamma \vdash \sigma \Rightarrow\Delta)\ra \idu\,\Delta\circ \sigma = \sigma
  Then, we show that the identity substitution itself is well typed:
  & \idw && : (\Gamma \vdash)\ra \Gamma \vdash \idu\,\Gamma \Rightarrow \Gamma

\subsubsection{Uniqueness of typing and uniqueness of types}
  We define what it means for a type $T$ (of the meta-theory) to be a proposition as follows:

@@

\subsubsection{UIP and functional extensionality}
None of these constructions use UIP. Functional extensionality is
 necessary because the metatheoretic or infinitary (untyped) $\Pi$ takes a metatheoretic function
 as an argument. An example of induction step that uses it lies in the proof
 that identity substitutions do not change types (this is an equation satisfied
 by a model of the QIIT), in particular in the case  $(\Pim\,T\,
 A)[\id]=\Pim\,T\,A$. Indeed, the left hand side of this equation is
 equal to $\Pim\,T\,(\lambda t.(A\,t)[\id])$ by definition, whereas the induction hypothesis
 states that $(t:T)\ra (A\,t)[\id]=A\,t$.
 % , which is not applyable without
 % functional extensionality.

\subsection{Functional relation enjoyed by initial morphism}
The previous section builds the syntactic model.
The challenge remains to show that it is initial.

Given a model $M$ of the QIIT, we define the functional relation enjoyed by the initial
morphism $\rec : S\to M$ by recursion on the typing judgments.
% As these judgments are propositions, we will sometimes
If $\Gamma$ is a context in $S$ and $\model{\Gamma}$ is a semantic
context (i.e. a context of the model $M$), we want to define a type $\Gamma\sim \model{\Gamma}$
equivalent to $\rec(\Gamma)=\model{\Gamma}$. Of course, at this
stage, $\rec$ is not available yet since the whole point of defining this
relation is to construct $\rec$ in the end. For a type $A$ in a context $\Gamma$,
it is natural to define the relation with a semantic type $\model{A}$ in a
semantic context $\model{\Gamma}$ such that it is equivalent to  $(\rec\,\Gamma =
\model{\Gamma})\times (\rec\,A = \model{A})$ (note that the first equality is
required for the second one to be well-typed).
The discussion is similar for terms.

The ultimate goal is to prove that
$\sum_{\model{\Gamma}}\Gamma\sim\model{\Gamma}$ and
$\sum_{\model{\Gamma}}({\model{A}:\model{\Ty}\,\model{\Gamma}})\times(A\sim\model{A})$
are contractible, i.e., have a unique inhabitant. The first step consists in showing that these types are propositions.
% We explain below that this requires UIP.
Actually, we take a different approach for the relation on types
(and terms) which yields a more concise definition of the relation.
The adapted goal, for types, will be to prove that
$\sum_{\model{A}}A\sim\model{A}$ is a proposition, and
is inhabited if if $\Gamma \sim \model{\Gamma}$ is.
 This means that in the definition of the relation, we assume
that contexts are already related, so that we don't need to enforce them to be,
as it is the case in the original approach. We give the two versions of the universe
case below, to explain the differences.

@@

In the definitions, we abbreviate $\wf{A} \sim (\model{\Gamma}\vdash \model{A})$ by
$\wf{A}\sim\model{A}$ when $\model{\Gamma}$ is clear, and similarly for terms.

@@

  & \blank \sim (\model{\Gamma}\vdash  \blank) && :  \Gamma \vdash A \, \ra \model{\Ty}\,\model{\Gamma} \ra \Set \\
  & \blank \sim{ (\model{\Gamma}\vdash  \blank \in \model{A})} && :  \Gamma \vdash t \in A \, \ra \model{\Tm}\,\model{\Gamma}\,\model{A} \ra \Set \\
  & \blank \sim{ (\model{\Gamma}\vdash  \blank \in \model{A})} && :  \Gamma \vdash x \invar A \, \ra \model{\Tm}\,\model{\Gamma}\,\model{A} \ra \Set \\
  & \blank \sim{ (\model{\Gamma}\vdash  \blank \Rightarrow \model{\Delta})} && :  \Gamma \vdash \sigma \Rightarrow \Delta \, \ra \model{\Sub}\,\model{\Gamma}\,\model{\Delta} \ra \Set \\

@@

  & \epsilonw\sim
      ( \model{\Gamma} \vdash \model{\delta}\Rightarrow \model{E})&&

@@

    & (\consw \wf{\Delta}\wf{\sigma}\wf{A}\wf{t})\sim
    & \qquad
    ( \model{\Gamma} \vdash \model{\delta}\Rightarrow \model{E})&&

>>>>

The steps consist of the following:
\begin{enumerate}
\item Definition of untyped syntax (as a family of inductive datatypes) together
  with typing judgments (as inductive relations on the untyped
  syntax), and construction of a model of the theory of IIT
  signatures from well-formed terms, denoted $S$.
\item Construction of a morphism $\rec:S\to M$ for arbitrary $M$, by:
  \begin{enumerate}
  \item defining a relation $\blank \sim \blank$ between the (well typed) syntax and a given
    model (the idea is that given a syntactic context $\Gamma$ and a semantic
    context $\model{\Gamma}$ of the model $M$, we have $\Gamma \sim
    \model{\Gamma}$ if and only if $\rec\,\Gamma = \model{\Gamma}$, and
    similarly for types, terms, and substitutions);
    % semantic context $\model{\Gamma}$ of
    % the model $M$ if and only if $\Gamma$ relates to $\rec\,{\Gamma}$;
    % enjoyed by the initial morphism from the
    % syntax to a given model;
  \item showing that this relation is functional.
    \end{enumerate}
  \item Proving of the uniqueness of this morphism by showing that
    any morphism $f:S\to M$ satisfies the relation
    (in the sense, for contexts, that given any syntactic context $\Gamma$, we
    have $\Gamma\sim f\,\Gamma$).
      % TODO donner composantes sur les contextes
\end{enumerate}
The next sections detail each of these steps.
\subsection{Syntactic model}

The goal is to define the syntactic model where contexts are pairs of an untyped 
context together with a proof that it is well formed, and similarly for types,
terms and substitutions.
The next sections are devoted to the definition of the untyped syntax and their associated
typing judgments.

@@

\begin{alignat*}{5}
  & \rlap{(1) Substitution calculus} \\[0.5em]
  & \Conu && : \Set \\
  & \Tyu  && : \Set \\
  & \Subu  && : \Set \\
  % & \Subpre  && : \Set \\
  & \Tmu  && : \Set \\
  & \emptyu && : \Conu \\
    & \epsilonu && : \Subu \\
  & \blank\extu\blank && : \Conu\ra\Tyu\ra\Conu \\
  & \blank\consu\blank && : \Subu\ra\Tmu\ra\Subu \\
  & \varu  && : \N \ra \Tmu \\
  & \rlap{(2) Sorts and operators} \\[0.5em]
  & \Uu && : \Tyu \\
  & \Elu && : \Tmu\ra\Tyu \\
  & \rlap{(3) Parameters} \\[0.5em]
  & \Piu && : \Tmu\ra\Tyu\ra\Tyu \\
  & \blank\appu\blank && : \Tmu\ra\Tmu\ra \Tmu \\
  & \rlap{(4-5) Metatheoretic and infinitary parameters} \\[0.5em]
  & \Pimu && : (T:\Set)\ra(T\ra\Tyu)\ra\Tyu \\
  & \Piiu && : (T:\Set)\ra(T\ra\Tmu)\ra\Tmu \\
  & \blank\appmu \blank && : \Tmu\ra(\alpha:T) \ra\Tmu\\
  & \rlap{(6) Default value} \\[0.5em]
  & \erru && : \Tmu
\end{alignat*}
Note that we choose to share the constructor for untyped application to infinitary and
meta-theoretic parameters.
Variables are modelled as De Bruijn indices, i.e. as natural numbers pointing
to the relevant position in the context.

@@

\subsubsection{Untyped weakening}

@@

Note that $(\Piu \,A\,B)[\sigma]$ should be defined as $\Piu
\,A[\sigma]\,B[\varu\, 0 \,\consu\, \wkO\,\sigma]$, and thus we need to define
$\wk0$, the weakening of substitutions.
The basic idea is to increment the de Bruijn indices of all the variables.
Actually this is not so simple because of the $\Piu$ type:
indeed, we want to define $\wk(\Piu\,A\,B)$ as the $\Pi$ type of the weakening
of $A$ and $B$, but here, $B$ must be weakened with respect to the second last
variable of the context, rather than the last one.
For this reason, we need to generalize the weakening as occuring anywhere in the context.
\begin{alignat*}{5}
  & \wk_n && :  \Tyu\ra\Tyu \\
  & \wk_n && :  \Tmu\ra\Tmu \\
  % & \wk && : \N \ra \N\ra\N \\
  & \wkO && : \Subu\ra \Subu
  \end{alignat*}
  The natural number $n$ specifies at which position of the context the
  weakening occurs.
  Here, $\wkO$ weakens with respect to the last variable.

  Later, in Section~\ref{ss:weaken-typing}, we show that weakening preserves typing.
  But as we just argued, weakening may occur anywhere in the context.
Stating a typing rule for this operation requires to have a way to express
the weakening occuring at the middle of a context.
To this end, we consider pairs of untyped contexts, which should be thought
of as a splitting at some place of a full context.
This full context is recovered by merging the two components:
% (which appears in the typing rule of a weakening occuring in the middle of a context),
% and the length $\length{\Gamma}$ of a
% context $\Gamma$:
\begin{alignat*}{5}
  & \blank \merge \blank && :  \Conu\ra\Conu\ra\Conu \\
  & \Gamma \merge \cdot && :=  \Gamma \\
  & \Gamma \merge (\Delta \extu A) && :=  (\Gamma \merge \Delta)\extu A
\end{alignat*}
We think of the second context as a telescope over the first context.
We need to define the define also the weakening of a telescope in order to
state the typing rule of weakening:
\begin{alignat*}{5}
  & \wkO && : \Conu\ra \Conu
  \\
  & \wkO\,\emptyu && := \emptyu
  \\
  & \wkO\,(\Delta \extu A) && := \wkO\,\Delta \extu \wk_{\length{\Delta}}\,A
  \end{alignat*}
  \subsubsection{Untyped substitution}
  We define by induction the specific substitution of one variable of the context:
  % & \rlap{(1) Substitution calculus} \\[0.5em]
\begin{alignat*}{5}
  & \blank[\blank := \blank] && : \Tyu\ra\N\ra\Tmu\ra\Tyu \\
  & \blank[\blank := \blank] && : \Tmu\ra\N\ra\Tmu\ra\Tmu
  % \\
  % & \blank[\blank := \blank] && : \N\ra\N\ra\Tmu\ra\Tmu
  \end{alignat*}
  This is enough to define the well-typed judgments: indeed, the typing rule of application
  involves only a unary substitution.

@@

\subsubsection{Weakening preserves typing}
\label{ss:weaken-typing}
We state and prove by mutual recursion that typing judgments are stable under
weakening, for contexts, types, terms, and substitutions:
% wkTelw : ∀ {Γp}{Ap}(Aw : Γp ⊢ Ap)Δp (Δw : (Γp ^^ Δp) ⊢) → ((Γp ▶p Ap) ^^ wkTel Δp) ⊢
% liftTw : ∀ {Γp}{Ap}(Aw : Γp ⊢ Ap)Δp{Bp}(Bw : (Γp ^^ Δp) ⊢ Bp) → ((Γp ▶p Ap) ^^ wkTel Δp) ⊢ (liftT ∣ Δp ∣ Bp)
\begin{alignat*}{5}
  & \wf{\wkO} && : (\Gamma \vdash A)\ra(\Gamma \merge \Delta\vdash) \ra
  \Gamma \extu A \merge \wkO\, \Delta \vdash \\
  % liftV (S (n + p)) (liftV n q) ≡ liftV n (liftV (n + p) q)
  & \wf{\wk} && : (\Gamma \vdash A)\ra(\Gamma \merge \Delta\vdash B) \ra
  \Gamma \extu A \merge \wkO\,\Delta \vdash \wk_{\length{\Delta}}\, B  \\
  & \wf{\wk} && : (\Gamma \vdash A)\ra(\Gamma \merge \Delta\vdash t\in B) \ra
  \Gamma \extu A \merge \wkO\,\Delta \vdash \wk_{\length{\Delta}}\, t \in \wk_{\length{\Delta}}\,B  \\
  & \wf{\wkO} && : (\Gamma \vdash A)\ra(\Gamma \vdash \sigma
  \Rightarrow \Delta) \ra
  \Gamma \extu A \vdash \wkO\, \sigma \Rightarrow \Delta
  \end{alignat*}
\subsubsection{Substitution preserves typing}
  We show that judgments are stable under substitution, by mutual recursion:
\begin{alignat*}{5}
  % Tyw[] : ∀ {Γp}{Ap}(Aw : Γp ⊢ Ap) {Δp}(Δw : Δp ⊢){σp}(σw :  Δp ⊢ σp ⇒ Γp) → Δp ⊢ (Ap [ σp ]T)
  & \wf{[]} && : (\Gamma \vdash)\ra(\Delta \vdash A)\ra
  (\Gamma \vdash \sigma \Rightarrow \Delta) \ra
  \Gamma \vdash A [ \sigma ]
  \\
  & \wf{[]} && : (\Gamma \vdash)\ra(\Delta \vdash t \in A)\ra
  (\Gamma \vdash \sigma \Rightarrow \Delta) \ra
  \Gamma \vdash t [\sigma] \in A [ \sigma ]
  \\
  & \wf{[]} && : (\Delta \vdash x \invar A)\ra
  (\Gamma \vdash \sigma \Rightarrow \Delta) \ra
  \Gamma \vdash x [\sigma] \in A [ \sigma ]
  % ∘w : ∀ {Γ} {Δ σ} (σw :  Δ ⊢ σ ⇒ Γ)
  %  {Y}(Yw : Y ⊢) {δ} (δw :  Y ⊢ δ ⇒ Δ) →
  %  Y ⊢  (σ ∘p δ) ⇒ Γ
  \\
  & \wf{{\circ}} && :
  (\Gamma \vdash) \ra
  (\Gamma \vdash \sigma \Rightarrow \Delta) \ra
  (\Delta \vdash \tau \Rightarrow E) \ra
  \Gamma \vdash \tau \circ \sigma \Rightarrow E
  % keepw : ∀ {Γp}(Γw : Γp ⊢){Δp}(Δw : Δp ⊢){σp}(σw :  Γp ⊢ σp ⇒ Δp) {Ap}(Aw : Δp ⊢ Ap ∈ Up ) → (Γp ▶p (Elp Ap [ σp ]T )) ⊢ (keep σp) ⇒ (Δp ▶p Elp Ap)
  % & \wf{keep-\El} && : (\Gamma \vdash)\ra(\Delta \vdash A)\ra
  % (\Gamma \vdash \sigma \Rightarrow \Delta) \ra
  % \Gamma \vdash A [ \sigma ]
  % \\
  \end{alignat*}
Then, we show that identity substitutions are neutral, for well-formed types,
terms, and substitutions:
\begin{alignat*}{5}
  % [idp]V : ∀ {Γ}{A}{x}(xw : Γ ⊢ x ∈v A) → (x [ idp ∣ Γ ∣ ]V) ≡ V x
  & [\idu] && : (\Gamma \vdash A)\ra A [ \idu\,\Gamma]  \\
  & [\idu] && : (\Gamma \vdash x \invar A)\ra x [ \idu\,\Gamma] = V x \\
  & [\idu] && : (\Gamma \vdash t \in A)\ra t [ \idu\,\Gamma] = t \\
  & \idru && : (\Gamma \vdash \sigma \Rightarrow\Delta)\ra \sigma \circ \idu\,\Gamma = \sigma
  \end{alignat*}
  In a seperate and independant recursion, one shows that:
  \begin{alignat*}{5}
  & \idlu && : (\Gamma \vdash \sigma \Rightarrow\Delta)\ra \idu\,\Delta\circ \sigma = \sigma
  \end{alignat*}
  Then, we show that the identity substitution itself is well typed:
  \begin{alignat*}{5}
  & \idw && : (\Gamma \vdash)\ra \Gamma \vdash \idu\,\Gamma \Rightarrow \Gamma
  \end{alignat*}

\subsubsection{Uniqueness of typing and uniqueness of types}
\label{ss:uniq-typing-types}
  We define what it means for a type $T$ (of the meta-theory) to be a proposition as follows:

@@

% \subsubsection{UIP and functional extensionality}
 % , which is not applyable without
 % functional extensionality.

 \subsubsection{Syntactic model}
 The syntactic model is obtained 
 by packing the untyped syntax
with the typing judgments:
% % \subsection{QIIT model}
% Then, we show that they are stable by weakening and substitution.
% Packing the untyped syntax with well-formed typed judgments yields
% a model $S$ of the QIIT. Let us give the most important components:
\begin{alignat*}{5}
 & \syn\Con && := \sum_\Gamma \Gamma\vdash
 \\
 & \syn\Ty \,(\Gamma,\wf\Gamma)&& := \sum_A \Gamma\vdash A
 \\
 & \syn\Tm \,(\Gamma,\wf\Gamma)(A,\wf A)&& := \sum_t \Gamma\vdash t \in A
 \\
 & \syn\Sub \,(\Gamma,\wf\Gamma)(\Delta,\wf \Delta)&& := \sum_\sigma \Gamma\vdash \sigma \Rightarrow \Delta
\end{alignat*}
The other fields are given straightforwardly.
Regarding the equations, it is enough to prove them only for the untyped syntactic
part: as we argued in Section~\ref{ss:uniq-typing-types}, the proofs of typing judgments are automatically
equal.

None of the constructions in this section uses UIP. Functional extensionality is
 necessary because the metatheoretic or infinitary (untyped) $\Pi$ takes a metatheoretic function
 as an argument. An example of induction step that uses it lies in the proof
 that identity substitutions do not change types (this is an equation satisfied
 by a model of the QIIT), in particular in the case  $(\Pim\,T\,
 A)[\id]=\Pim\,T\,A$. Indeed, the left hand side of this equation is
 equal to $\Pim\,T\,(\lambda t.(A\,t)[\id])$ by definition, whereas the induction hypothesis
 states that $(t:T)\ra (A\,t)[\id]=A\,t$.

\subsection{A functional relation}
The previous section builds the syntactic model.
The challenge remains to show that it is initial.

Given a model $M$ of the QIIT, we define the functional relation satisfied by the initial
morphism $\rec : S\to M$ by recursion on the typing judgments.
% As these judgments are propositions, we will sometimes
If $\Gamma$ is a context in $S$ and $\model{\Gamma}$ is a semantic
context (i.e. a context of the model $M$), we want to define a type $\Gamma\sim \model{\Gamma}$
equivalent to $\rec(\Gamma)=\model{\Gamma}$. Of course, at this
stage, $\rec$ is not available yet since the whole point of defining this
relation is to construct $\rec$ in the end.

For a type $A$ in a context $\Gamma$, we want to define a relation
$A\sim \model{A}$ that is equivalent to $\rec(A)=\model{A}$.
% where $\model{A}$ is a
% semantic type in a semantic context $\model{\Gamma}$.
For this equality to make sense, the semantic type $\model{A}$ must live
in the semantic context $\rec\ \Gamma$. But again, $\rec$ is not
yet available at this stage. Exploiting the expected equivalence between
$\Gamma\sim\model\Gamma$ and $\rec(\Gamma)=\model\Gamma$, we may consider defining $A\sim\model{A}$ under the
hypotheses that $\model{A}$ lies in a semantics context $\model\Gamma$ which is
related to $\Gamma$. Then, the type of the relation for types is
\[
  (\Gamma : \syn\Con)\ra (A : \syn\Ty\syn\Gamma) \ra
  (\model\Gamma : \model\Con)
  \ra
  (\Gamma \sim \model\Gamma)
  \ra
  (\model{A}:\model\Ty\model\Gamma)
  \ra
  \Set
\]
Note that the relation on contexts
must be defined mutually with the relation on types (see for example the case of
context extension), but here, the relation on contexts appears as the type of an
argument of the relation on types.
% We leave this possibility aside,
As we do not assume that our metatheory
supports this kind of recursive-recursive definitions, we refuse to further consider
 such a possiblity, and instead just remove the 
hypothesis $\Gamma\sim\model\Gamma$ from the list of arguments.
We proceed similarly for terms and substitutions. Actually, 
this removal is not without harm. Consider for example the case of relating the
empty substitution $\Gamma\vdash \epsilonu \Rightarrow \emptyu $
with a semantic substitution $\model\sigma$ between semantic contexts
$\model\Gamma$ and $\model\Delta$. We would like to assert that $\model\sigma$
equals the empty semantic substitution $\model\epsilon$, but this is not
possible because typechecking requires that $\model\Delta$ is 
the empty semantic context. This is precisely what ensured the removed hypothesis
$\syn\cdot \sim \model\Delta$.
Our way out here is to state that $\model\sigma$ relates to the empty
substitution if the target semantic context is empty, and, acknowledging
this equality, if $\model\sigma$ is
the empty substitution.
% The idea is that we will first prove that there is atmost one semantic type
% which relates to the syntactic one, and then later we will provide such a
% semantic type under the hypothesis that the contexts are related.

Let us mention another possible solution avoiding the recursive-recursive
requirement: it consists in defining $A\sim\model{A}$ so that it is equivalent to
$(e:\rec\,\Gamma = \model{\Gamma})\times (\rec\,A = e\#\model{A})$.
% whereas the right hand side is of type $\model\Ty\model{\Gamma}$.
% For a substitution $\sigma$ between contexts $\Gamma$ and $\Delta$,
% a first idea consists in defining a relation $\sigma \sim_{\blank} \model{\sigma} : \Gamma
% \sim \model\Gamma \to \Set $ where $\model{A}$ is a semantic type in the
% semantic context $\model{\Gamma}$. We define the relation on types under the hypothesis
% that the contexts are already related.
% it is natural to define the relation with a semantic type $\model{A}$ in a
% semantic context $\model{\Gamma}$ such that it is equivalent to  $(\rec\,\Gamma = \model{\Gamma})\times (\rec\,A = \model{A})$ (note that the first equality is
% required for the second one to be well-typed).
% The discussion is similar for terms.
% Then, the ultimate goal is to prove that
% $\sum_{\model{\Gamma}}\Gamma\sim\model{\Gamma}$ and
% $\sum_{\model{\Gamma}}({\model{A}:\model{\Ty}\,\model{\Gamma}})\times(A\sim\model{A})$
% are contractible, i.e., have a unique inhabitant.
%The first step consists in showing that these types are propositions.
Our different approach
% Actually, we take a different approach for the relation on types
% (and terms) which
yields a more concise definition of the relation.
For example, in the case of the universe, this would lead
to the definition
  $  \Uw\wf{\Gamma} \sim \model{A} := (\wf{\Gamma}\sim\model{\Gamma}) \times (\model{A} = \model{\U})$,
  instead of
  our definition
  $  \Uw\wf{\Gamma} \sim \model{A} := (\model{A} = \model{\U})$.
% Recall that our adapted goal, for types, is to prove that
% $\sum_{\model{A}}A\sim\model{A}$ is a proposition, and
% is inhabited if $\Gamma \sim \model{\Gamma}$ is.
%  This means that in the definition of the relation, we assume
% that contexts are already related, so that we don't need to enforce them to be,
% as it is the case in the original approach.
% We give the two versions of the universe
% case below, to explain the differences.

@@

In the definitions, we abbreviate $\relT{\wf{A}}{\model\Gamma}{\model{A}}$ by
$\wf{A}\sim\model{A}$ when $\model{\Gamma}$ can be inferred, and similarly for
terms and substitutions.

@@

  & \relT{\blank}{\model{\Gamma}}{\blank} && :  \Gamma \vdash A \, \ra \model{\Ty}\,\model{\Gamma} \ra \Set \\
  & \relt{\blank}{\model{\Gamma}}{\model{A}}{\blank} && :  \Gamma \vdash t \in A \, \ra \model{\Tm}\,\model{\Gamma}\,\model{A} \ra \Set \\
  & \relV{\blank}{\model{\Gamma}}{\model{A}}{\blank}&& :  \Gamma \vdash x \invar A \, \ra \model{\Tm}\,\model{\Gamma}\,\model{A} \ra \Set \\
  & \relS{\blank}{\model{\Gamma}}{\model{\Delta}}{ \blank} && :  \Gamma \vdash \sigma \Rightarrow \Delta \, \ra \model{\Sub}\,\model{\Gamma}\,\model{\Delta} \ra \Set \\

@@

  & \relS{\epsilonw}{\model{\Gamma}}{\model{E}}{\model{\delta}}
  % (  \vdash \model{\delta}\Rightarrow \model{E})

@@

    &
    \relS{(\consw \wf{\Delta}\wf{\sigma}\wf{A}\wf{t})}{\model{\Gamma}}{\model{E}}{\model{\delta}}
    % (\consw \wf{\Delta}\wf{\sigma}\wf{A}\wf{t})\sim
    % & \qquad
    % ( \model{\Gamma} \vdash \model{\delta}\Rightarrow \model{E})
